---
title: "Matrix and Linear Models Exercises"
author: "Emre Usenmez - github.com/emre-us"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# HarvardX PH525.2x: Intro to Linear Models and Matrix Algebra - Exercises

## Intro Exercises

Install the library UsingR:

```{r}
install.packages("UsingR")
```

Access Galton's father and son heights:

```{r}
library(UsingR)
data("father.son", package = "UsingR")
```

Q1. What is the average height of the sons?

```{r}
summary(father.son)
```

One of the defining features of regression is that we stratify one variable based on others. In Statistics we use the verb "condition". For example, the linear model for son and father heights answers the question how tall do I expect a son to be if I condition on his father being x inches. The regression line answers this question for any x.

Using Galton's dataset, we want to know the expected height of sons if we condition on the father being 71 inches. Create a list of son heights for sons that have fathers with heights of 71 inches (round to the nearest inch). What is the mean of the son heights for fathers that have a height of 71 inches?

```{r}
library(dplyr)
fh71 <- father.son %>%
  filter(round(fheight) == 71)

mean(fh71$sheight)

# Alternatively:
mean(father.son$sheight[round(father.son$fheight)==71])
```










## Matrix Notation Exercises

In R we have vectors and matrices. You can create your own vectors with the function c(). Vectors are also the output of many functions such as:

```{r}
rnorm(5)
```

You can turn vectors into matrices using functions such as rbind(), cbind(), or matrix().

```{r}
X = matrix(1:1000, 100, 10)
```

Q1. What is the entry in row 25, column 3?

```{r}
X[25,3]
```

Q2. Using the function cbind(), create a 10x5 matrix with first column x=1:10. Then use 2*x, 3*x, 4*x, and 5*x, respectively in columns 2 through 5. What is the sum of the elements of the 7th row?

```{r}
x1 <- 1:10

X <- cbind(x1, x*2, x*3, x*4, x*5)
dim(X)

sum(X[7,])
```

Q3. Create a matrix with multiples in the 3rd column.

```{r}
X <- matrix(1:60,20,3, byrow = TRUE)

#to check if 3rd column has all multiples of 3:
all(X[,3]%%3==0) #The symbol "%%" refer to an arithmetic operation called modulo operation. It gives the remainder after x divided by y. So here if all the 3rd column is divided by 3, the remainder would need to be 0 for the entire column to be multiples of 3. This is a quick way of checking it.

```

Q5. What is the last element of the vector returned by seq(10, 1, -2)?

```{r}
seq(10,1,-2)
```










## Notes on Inverse Matrix

The inverse of matrix X, denoted with $X^-1$ has the property that, when multiplied, gives you the identity $X^{-1}X=I$. 

This is an identity matrix:
```{r}
diag(4)
```


Of course, not all matrices have inverses. For example, a $2\times 2$ matrix with 1s in all its entries does not have an inverse. 

As we will see when we get to the section on applications to linear models, being able to compute the inverse of a matrix is quite useful. A very convenient aspect of R is that it includes a predefined function `solve` to do this. Here is how we would use it to solve the linear of equations:

$$
\begin{align*}
a + b + c &= 6\\
3a - 2b + c &= 2\\
2a + b  - c &= 1
\end{align*}
$$
We described how this system can be rewritten and solved using matrix algebra:

$$
\,
\begin{pmatrix}
1&1&1\\
3&-2&1\\
2&1&-1
\end{pmatrix}
\begin{pmatrix}
a\\
b\\
c
\end{pmatrix} =
\begin{pmatrix}
6\\
2\\
1
\end{pmatrix}
\implies
\begin{pmatrix}
a\\
b\\
c
\end{pmatrix}=
\begin{pmatrix}
1&1&1\\
3&-2&1\\
2&1&-1
\end{pmatrix}^{-1}
\begin{pmatrix}
6\\
2\\
1
\end{pmatrix}
$$
Here is how we would use it to solve the linear of equations in R:

```{r}
X <- matrix(c(1,3,2,1,-2,1,1,1,-1),
            3,3)
y <- matrix(c(6,2,1), 
            3, 1)

solve(X)%*%y #equivalent to solve (X,y). "%*% is to do a matrix multiplication. Note that "solve()" is a function that should be used with caution as it is not generally numerically stable.
```










## Matrix Operation Exercises

Q2. Solve for the following system of equations:

$$
\begin{align*}
3a + 4b + 5c + d &= 10\\
2a + 2b + 2c - d &= 5\\
a - b  + 5c - 5d &= 7\\
5a + d &= 4\\
\end{align*}
$$
```{r}
X <- matrix(c(3, 2, 1, 5, 4, 2, -1, 0, -5, 2, 5, 0, 1, -1, -5, 1), 
            4, 4)
y <- matrix(c(10, 5, 7, 4))

solve(X,y)
```


Load the following two matrices into R:

```{r}
a <- matrix(1:12, nrow = 4)
b <- matrix(1:15, nrow = 3)
```

In the questions below we will use the matrix multiplication operator in R, $ %*% $, to multiply these two matrices.

Q3. What is the value in the 3rd row and 2nd column of the matrix product of a and b?

```{r}
a%*%b
```

Q4. Multiply the 3rd row of a with the second column of b using the element-wise vector multiplication with *. What is the sum of the elements in the resulting vector?

```{r}
a[3,] %*% b[,2]

# alternatively

sum(a[3,]*b[,2])
```

This is equivalent to the 3rd row, 2nd column element of the product of the two matrices.
